Item Construction Rules Revisited: Learnings from Measurement of Latent Variables with Gold Standard Items

### 2025-07-17 manuscript.qmd
manual date handling

---
title: "Item Construction Rules Revisited: Learnings from Measurement of Latent Variables with Gold Standard Items"
shorttitle: "Measuring latent variables with gold standard items"
date: today
date-format: iso
author:
  - name: Kathryn Eichhorn
    equal-contributor: true
    orcid: 0000-0003-3676-9420
    role: 
      - conceptualization
      - investigation
      - resources
      - data curation
      - writing
    affiliations:
      - name: University of the Bundeswehr Munich
        department: Institute of Psychology
        city: Neubiberg
        country: Germany
  - name: Markus Bühner
    corresponding: true
    equal-contributor: true
    email: buehner@lmu.de
    orcid: 0000-0002-0597-8708
    role: 
      - formal analysis
      - supervision
      - writing
    affiliations:
      - id: "lmu"
        name: LMU Munich
        department: Department of Psychology
        address: Leopoldstr. 13
        city: D-80802 Munich
        country: Germany
        postal-code: 80802
  - name: Florian Pargent
    orcid: 0000-0002-2388-553X
    role:
      - formal analysis
      - software
      - validation
      - editing
    affiliations:
      - ref: lmu
  - name: Janika Saretzki
    orcid: 0000-0002-6536-8266
    role:
      - editing
    affiliations:
      - name: University of Graz
        department: Department of Psychology
        city: Graz
        country: Austria
      - name: Charlotte Fresenius University of Psychology
        city: Munich
        country: Germany
      - ref: lmu
  - name: Larissa Sust
    orcid: 0000-0002-3389-1626
    role:
      - editing
    affiliations:
      - ref: lmu
  - name: Jonas Hauck
    orcid: 0009-0002-8872-0530
    role:
      - editing
    affiliations:
      - id: "regen"
        name: University of Regensburg
        department: Faculty of Psychology
        city: Regensburg
        country: Germany
  - name: Sven Hilbert
    orcid: 0000-0001-5808-8357
    role:
      - conceptualization
      - supervision
      - editing
    affiliations:
      - ref: regen

author-note:
  disclosures:
    data-sharing: This is version 1 (last modification {{< meta date >}}) of our preprint published on PsyArXiv. All materials (reproducible manuscript, analysis code, datasets, codebooks, questionnaires) are available in the project's repository on the Open Science Framework (OSF) at <https://osf.io/p7492/>. A Quarto Manuscripts website is hosted at <https://florianpargent.github.io/gold-standard-items/>. The data has previously been analyzed in the dissertation thesis of the first author at <https://doi.org/10.5282/edoc.23628>. As part of the dissertation, a preregistration was published at <https://osf.io/cz3uv/?view_only=9cac02db231b48629fea9ae53c3038b9> that differs from the analyses reported in the current manuscript.
    conflict-of-interest: The authors declare that there were no conflicts of interest concerning the authorship or the publication of this article.
    financial-support: ~
    gratitude: ~
    authorship-agreements: ~
language:
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; <https://credit.niso.org/>):"
abstract: "This article investigates whether gold-standard items are helpful for questionnaire construction. Based on cognitive interviews (n = `r nrow(dat_interviews)`) and a pilot study (n = `r nrow(dat_pilot)`) with an initial item set, three 12-item scales were constructed to measure the physical traits of body height, body weight, and age. We collected data on these scales using response formats with either two (n = `r nrow(dat_dich)`) or six (n = `r nrow(dat_poly)`) categories. We also collected numeric self-reports of body weight, body height, and age as gold-standard items. Confirmatory factor analyses revealed that the gold-standard items did not consistently exhibit the highest loading on their corresponding latent variable. Furthermore, when controlling for the self-reported physical body height, body weight, and age as gold-standard items, as well as gender, we did not always find an interpretable, systematic residual variance. Finally, the pattern of correlations between the latent variables did not reflect the correlations between the self-reported gold-standard items, suggesting that the item scales and the gold-standard items do not have the same validity. While these results are consistent with previous studies, our analyses also showed that items with two response categories were at least as valid as those with six categories, contradicting past recommendations. When constructing a questionnaire, we would argue that the items intended to measure the latent variable most directly should have the highest loading on that variable. If this is not the case, the content validity is questionable at best. The implications are that intensive cognitive pretesting is necessary. Questionnaires with different response formats should be compared empirically and hypotheses about which items best represent the latent variable should be tested."
license: CC BY 4.0
bibliography: [references.bib, r-references.bib]
floatsintext: true
numbered-lines: false
csl: apa.csl
---

>>>>

---
title: "Item Construction Rules Revisited: Learnings from Measurement of Latent Variables with Gold Standard Items"
shorttitle: "Measuring latent variables with gold standard items"
date: 2025-07-17
date-modified: 2025-07-17
date-format: iso
author:
  - name: Kathryn Eichhorn
    equal-contributor: true
    orcid: 0000-0003-3676-9420
    role: 
      - conceptualization
      - investigation
      - resources
      - data curation
      - writing
    affiliations:
      - name: University of the Bundeswehr Munich
        department: Institute of Psychology
        city: Neubiberg
        country: Germany
  - name: Markus Bühner
    corresponding: true
    equal-contributor: true
    email: buehner@lmu.de
    orcid: 0000-0002-0597-8708
    role: 
      - formal analysis
      - supervision
      - writing
    affiliations:
      - id: "lmu"
        name: LMU Munich
        department: Department of Psychology
        address: Leopoldstr. 13
        city: D-80802 Munich
        country: Germany
        postal-code: 80802
  - name: Florian Pargent
    orcid: 0000-0002-2388-553X
    role:
      - formal analysis
      - software
      - validation
      - editing
    affiliations:
      - ref: lmu
  - name: Janika Saretzki
    orcid: 0000-0002-6536-8266
    role:
      - editing
    affiliations:
      - name: University of Graz
        department: Department of Psychology
        city: Graz
        country: Austria
      - name: Charlotte Fresenius University of Psychology
        city: Munich
        country: Germany
      - ref: lmu
  - name: Larissa Sust
    orcid: 0000-0002-3389-1626
    role:
      - editing
    affiliations:
      - ref: lmu
  - name: Jonas Hauck
    orcid: 0009-0002-8872-0530
    role:
      - editing
    affiliations:
      - id: "regen"
        name: University of Regensburg
        department: Faculty of Psychology
        city: Regensburg
        country: Germany
  - name: Sven Hilbert
    orcid: 0000-0001-5808-8357
    role:
      - conceptualization
      - supervision
      - editing
    affiliations:
      - ref: regen

author-note:
  disclosures:
    data-sharing: This is version 1 (last modification {{< meta date-modified >}}) of our preprint published on PsyArXiv. All materials (reproducible manuscript, analysis code, datasets, codebooks, questionnaires) are available in the project's repository on the Open Science Framework (OSF) at <https://osf.io/p7492/>. A Quarto Manuscripts website is hosted at <https://florianpargent.github.io/gold-standard-items/>. The data has previously been analyzed in the dissertation thesis of the first author at <https://doi.org/10.5282/edoc.23628>. As part of the dissertation, a preregistration was published at <https://osf.io/cz3uv/?view_only=9cac02db231b48629fea9ae53c3038b9> that differs from the analyses reported in the current manuscript.
    conflict-of-interest: The authors declare that there were no conflicts of interest concerning the authorship or the publication of this article.
    financial-support: ~
    gratitude: ~
    authorship-agreements: ~
language:
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; <https://credit.niso.org/>):"
abstract: "This article investigates whether gold-standard items are helpful for questionnaire construction. Based on cognitive interviews (n = `r nrow(dat_interviews)`) and a pilot study (n = `r nrow(dat_pilot)`) with an initial item set, three 12-item scales were constructed to measure the physical traits of body height, body weight, and age. We collected data on these scales using response formats with either two (n = `r nrow(dat_dich)`) or six (n = `r nrow(dat_poly)`) categories. We also collected numeric self-reports of body weight, body height, and age as gold-standard items. Confirmatory factor analyses revealed that the gold-standard items did not consistently exhibit the highest loading on their corresponding latent variable. Furthermore, when controlling for the self-reported physical body height, body weight, and age as gold-standard items, as well as gender, we did not always find an interpretable, systematic residual variance. Finally, the pattern of correlations between the latent variables did not reflect the correlations between the self-reported gold-standard items, suggesting that the item scales and the gold-standard items do not have the same validity. While these results are consistent with previous studies, our analyses also showed that items with two response categories were at least as valid as those with six categories, contradicting past recommendations. When constructing a questionnaire, we would argue that the items intended to measure the latent variable most directly should have the highest loading on that variable. If this is not the case, the content validity is questionable at best. The implications are that intensive cognitive pretesting is necessary. Questionnaires with different response formats should be compared empirically and hypotheses about which items best represent the latent variable should be tested."
license: CC BY 4.0
bibliography: [references.bib, r-references.bib]
floatsintext: true
numbered-lines: false
csl: apa.csl
---
